### 2020-06-11 report/sections/server/extensions/async.tex
Report: complete section 6.1 on async

\subsection{Motivation}

@@

\subsection{API Extension}
\begin{itemize}
\item should be very short section -- use union type MaybePromise so developer can provide either
\end{itemize}

@@

\begin{lstlisting}[language=javascript,tabsize=2]
new Implementation.S40({
	Login: (*@\hl{async}@*) (username: string, password: string) => {
		const result = (*@\hl{await}@*) db.lookup(username, password);
		if (result.valid)
			return new Implementation.S42('Authorised', ...);
		else if (result.unauthorised)
			return new Implementation.S42('AccessDenied', ...);
		else
			return new Implementation.S42('Challenge', ..., challenger);
	}
});

const challenger = (key: number) => {
	return new (*@\hl{Promise}@*)((resolve, reject) => {
		keychain.check(key, ok => {
			if (ok)
				resolve(new Implementation.S44('Authorised', ...);
			else 
				resolve(new Implementation.S44('AccessDenied', ...);	
		});
	});
}
\end{lstlisting}

\subsection{Runtime Extension}

Generally, minimal changes are required for the runtime.
The key observation is that the concrete types for the any
handlers are the same -- they just might be wrapped in a Promise
construct.
To handle potential Promises, we wrap existing logic
as a delayed computation, colloquially known as a \textit{thunk}
\cite{Thunk}.

\begin{itemize}
\item server send actions -- use the Thunk pattern to wrap the existing logic, the thunk takes the original payload type (extracted using conditional infer types) as parameter; implementation either returns promise or not-promise, so the implementation will either pass thunk as resolve or call thunk as-is
\item server receive actions - thunk pattern as well
\item browser send action -- need a runtime flag to silently disable channel reuse (which would otherwise violate linearity) because the sending action is delayed by the Promise resolve
\item browser receive action -- thunk pattern
\end{itemize}

\begin{figure}[!h]
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript,tabsize=2]
const [	label, payload, succ] = this.handler;
send(label, payload);
next(succ);
\end{lstlisting}
\caption{\dots}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript,tabsize=2]
const thunk = (label, payload, succ) => {
	send(label, payload);
	next(succ);	
};

if (this.handler instanceof Promise)
	this.handler.then(thunk);
else 
	thunk(this.handler);
\end{lstlisting}
\caption{\dots}
\end{subfigure}
\captionof{figure}{\dots}
\end{figure}

\begin{lstlisting}[language=javascript,tabsize=2]
(*@\hl{let used = false;}@*)
...
return class extends React.Component {
	render() {
		const props = {
			[eventLabel as string]: (event) => {
				(*@\hl{if (used) return;}@*)
				(*@\hl{used = true;}@*)

				const result = handler(event);
				if (result instanceof Promise) {
					result.then(send);				
				} else {
					send(result);				
				}
			}		
		}
	}
}
\end{lstlisting}

\subsection{Limitations}

Supporting asynchronous logic on top of our
existing implementation in \fancyname{ReactMPST} comes at a cost
of losing affine linearity guarantees.
If the send action handler is asynchronous, then the UI event handler
returns before the runtime transitions, so the UI event that can trigger
the send action is still active, which means the user can trigger
a channel linearity violation. This means we need a boolean flag
to keep track of channel usage for browser-side endpoints, 
as done in \cite{Hybrid2016}.

>>>>

Asynchronous APIs are commonplace in modern web programming,
ranging from database operations to third-party API calls.
We motivate the need for supporting asynchronous operations
through an authentication protocol in
\cref{subsection:asyncmotivation},
demonstrate how we support this through extending
the generated APIs and runtime in
\cref{subsection:asyncapi,subsection:asyncruntime}
respectively, 
and highlight the limitations of our approach in
\cref{subsection:asynclimit}.

\subsection{Motivation}
\label{subsection:asyncmotivation}

@@

\subsection{API Extension}
\label{subsection:asyncapi}

This is rather straightforward.
We construct a generic union type, \texttt{MaybePromise<T>},
to specify that the type parameter might be wrapped in a Promise,
and leverage type inference in conditional types to
define a generic type, \texttt{FromPromise<T>}, to extract
the wrapped type.

@@

Now we change the type signatures of our generated APIs
to use \texttt{MaybePromise<T>}.
For example, if the handler of a receive state generated by
\fancyname{NodeMPST} must return \texttt{Implementation.S42},
it is now permitted to return a \texttt{Promise} that resolves
to a value of type texttt{Implementation.S42}.
The same logic applies to other types of EFSM states
in both \fancyname{NodeMPST} and \fancyname{ReactMPST}.

Hence, the asynchronous \trole{Svr} implementation
for the 2FA protocol shown in \cref{lst:new2fa}
now type-checks. Note that, because 
\lstonelinejs{async} functions return \texttt{Promise}s,
we also support this newer piece of syntactic sugar.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
new Implementation.S40({
	Login: (*@\hl{async}@*) (username: string, password: string) => {
		const result = (*@\hl{await}@*) db.lookup(username, password);
		if (result.valid)
			return new Implementation.S42('Authorised', ...);
		else if (result.unauthorised)
			return new Implementation.S42('AccessDenied', ...);
		else
			return new Implementation.S42('Challenge', ..., challenger);
	}
});

const challenger = (key: number) => {
	return new (*@\hl{Promise}@*)((resolve, reject) => {
		keychain.check(key, ok => {
			if (ok)
				resolve(new Implementation.S44('Authorised', ...);
			else 
				resolve(new Implementation.S44('AccessDenied', ...);	
		});
	});
}
\end{lstlisting}
\captionof{lstlisting}{\tprotocol{2FA} \trole{Svr} Implementation using 
Callbacks, \texttt{Promise}s and
\lstonelinejs{async}/\lstonelinejs{await}}
\label{lst:new2fa}
\end{figure}

\subsection{Runtime Extension}
\label{subsection:asyncruntime}

Generally, minimal changes are required for the runtime.
The key observation is that the concrete types for the any
handlers are the same -- they just might be wrapped in a Promise
construct.
To handle potential Promises, we wrap existing logic
as a delayed computation, colloquially known as a \textit{thunk}
\cite{Thunk}.
We show an example in \cref{fig:thunk} of how we
adapt the \texttt{performSend()} function generated by
\fancyname{NodeMPST} to support optional \texttt{Promise}s:

\begin{enumerate}
\item
We wrap the original logic in \cref{subfig:originalthunk} into a
function (\cref{subfig:newthunk} \cref{line:thunk})
to delay the computation.

\item
We distinguish asynchronous implementations
from synchronous ones using the \lstonelinejs{instanceof} operator.

\begin{itemize}
\item
If the implementation is a \texttt{Promise}, we pass
the thunk as the resolve function invoked on completion.

\item
Otherwise, we directly invoke the computation.
\end{itemize}

\end{enumerate}

\begin{figure}[!h]
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript,tabsize=2]
const [label, payload, succ] = this.handler;
send(label, payload);
next(succ);
\end{lstlisting}
\caption{Original \texttt{performSend()} Implementation}
\label{subfig:originalthunk}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
\begin{lstlisting}[language=javascript,tabsize=2]
const thunk = (label, payload, succ) => { (*@\label{line:thunk}@*)
	send(label, payload);
	next(succ);	
};

if (this.handler instanceof Promise)
	this.handler.then(thunk);
else
	thunk(this.handler);
\end{lstlisting}
\caption{New \texttt{performSend()} for Optional \texttt{Promise}s}
\label{subfig:newthunk}
\end{subfigure}
\caption{Adapting Send Operations in \fancyname{NodeMPST}
using a Thunk}
\label{fig:thunk}
\end{figure}

Similar changes apply to other types of states in both
\fancyname{NodeMPST} and \fancyname{ReactMPST}.
For \fancyname{ReactMPST}, channel-related functions in the runtime
return the successor state identifier. Hence, aside from
refactoring the functions to delay computation using thunks,
the functions themselves need to
return \texttt{MaybePromise<State>} instead
to propagate the asynchrony
to the execution of the EFSM -- the transition occurs
\textit{after} the developer's handler finishes execution.

\subsection{Limitations}
\label{subsection:asynclimit}

Supporting asynchronous logic on top of our
existing implementation in \fancyname{ReactMPST} comes at a cost
of losing affine linearity guarantees.
If the send action handler is asynchronous, then the UI event handler
returns before the runtime transitions, so the UI event that can trigger
the send action is still active, which means the user can trigger
a channel linearity violation. This means we need a boolean flag
to keep track of channel usage, as done in \cite{Hybrid2016}.
We illustrate this in \cref{lst:asynclinearcheck},
with the linearity checks highlighted.

\begin{figure}[!h]
\begin{lstlisting}[language=javascript,tabsize=2]
(*@\hl{let used = false;}@*)
...
return class extends React.Component {
	render() {
		const props = {
			[eventLabel as string]: (event) => {
				(*@\hl{if (used) return;}@*)
				(*@\hl{used = true;}@*)

				const result = handler(event);
				if (result instanceof Promise)
					result.then(send);
				else
					send(result);
			}		
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{\dots}
\label{lst:asynclinearcheck}
\end{figure}
